---
title: "DESeq2 differential expression via rosetta-bioc"
subtitle: "Posit Cloud demo — Python + R + Bioconductor"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(reticulate)
use_python(Sys.which("python3"), required = TRUE)
knitr::knit_engines$set(python = reticulate::eng_python)
```

## Prerequisites

```{r install, message=TRUE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("DESeq2", quietly = TRUE))
    BiocManager::install("DESeq2", ask = FALSE)
```

```{bash pip-install, eval=FALSE}
pip install --quiet rosetta-bioc matplotlib
```

---

## 1. Simulate a count dataset

We simulate 500 genes × 6 samples (3 treated, 3 untreated) with ~200 truly
differentially expressed genes. This keeps memory well within Posit Cloud's
limits while exercising the full rosetta + DESeq2 pipeline.

```{r load-airway}
set.seed(42)
n_genes   <- 500
n_samples <- 6

# baseline counts drawn from negative binomial
baseline <- rnbinom(n_genes * n_samples, mu = 100, size = 5)
counts_r  <- matrix(baseline, nrow = n_genes, ncol = n_samples)

# inject 200 DE genes: 2× up in treated samples
de_idx <- 1:200
counts_r[de_idx, 4:6] <- round(counts_r[de_idx, 4:6] * 2)

rownames(counts_r) <- paste0("gene", seq_len(n_genes))
colnames(counts_r) <- paste0("sample", seq_len(n_samples))

coldata_r <- data.frame(
    dex = c("untrt", "untrt", "untrt", "trt", "trt", "trt"),
    row.names = colnames(counts_r)
)

cat("Genes:  ", nrow(counts_r), "\n")
cat("Samples:", ncol(counts_r), "\n")
print(table(coldata_r$dex))
```

```{python hand-to-python}
import numpy as np
import pandas as pd
import rpy2.robjects as ro
from rpy2.robjects import pandas2ri

r_counts  = ro.r["counts_r"]
r_coldata = ro.r["coldata_r"]

genes   = list(r_counts.rownames)
samples = list(r_counts.colnames)

# rpy2 flattens R matrices column-major; reshape then transpose → (genes × samples)
arr = np.array(r_counts).reshape(len(genes), len(samples), order="F")
counts = pd.DataFrame(arr, index=genes, columns=samples)

with (ro.default_converter + pandas2ri.converter).context():
    metadata = ro.conversion.rpy2py(r_coldata)

metadata.index  = samples
metadata["dex"] = metadata["dex"].astype(str)

print(f"Count matrix : {counts.shape[0]:,} genes × {counts.shape[1]} samples")
print(f"Conditions   : {metadata['dex'].value_counts().to_dict()}")
counts.iloc[:3, :4]
```

---

## 2. Run DESeq2 via rosetta

One call — `quick_deseq2` builds the DESeqDataSet, fits the model, and returns
results as a **pandas DataFrame**.

```{python run-deseq2}
import rosetta as rb

results = rb.quick_deseq2(
    counts,
    metadata,
    design="~ dex",
    alpha=0.05,
)

print(type(results))
results.head()
```

---

## 4. Summary

```{python summary}
results.report()
```

---

## 5. Top differentially expressed genes

```{python top-genes}
sig = results.dropna(subset=["padj"]).query("padj < 0.05")
print(f"{len(sig):,} significant genes (FDR < 5%)")

top10 = sig.sort_values("log2FoldChange", key=abs, ascending=False).head(10)
top10[["baseMean", "log2FoldChange", "lfcSE", "padj"]]
```

---

## 6. Volcano plot

```{python volcano, fig.width=8, fig.height=6}
import matplotlib
matplotlib.use("Agg")
import matplotlib.pyplot as plt

plot_df = results.dropna(subset=["padj", "log2FoldChange"]).copy()
plot_df["neg_log10_padj"] = -np.log10(plot_df["padj"].clip(lower=1e-300))

is_sig = plot_df["padj"] < 0.05
is_up  = is_sig & (plot_df["log2FoldChange"] >  1)
is_dn  = is_sig & (plot_df["log2FoldChange"] < -1)

fig, ax = plt.subplots(figsize=(8, 6))

ax.scatter(plot_df.loc[~is_sig, "log2FoldChange"],
           plot_df.loc[~is_sig, "neg_log10_padj"],
           s=6, color="#aaaaaa", alpha=0.5, label="not significant")
ax.scatter(plot_df.loc[is_up, "log2FoldChange"],
           plot_df.loc[is_up, "neg_log10_padj"],
           s=10, color="#e05c5c", alpha=0.8, label=f"up ({is_up.sum()})")
ax.scatter(plot_df.loc[is_dn, "log2FoldChange"],
           plot_df.loc[is_dn, "neg_log10_padj"],
           s=10, color="#5c88e0", alpha=0.8, label=f"down ({is_dn.sum()})")

ax.axhline(-np.log10(0.05), color="black", lw=0.8, ls="--", alpha=0.6)
ax.axvline( 1, color="black", lw=0.8, ls="--", alpha=0.4)
ax.axvline(-1, color="black", lw=0.8, ls="--", alpha=0.4)

for gene, row in top10.head(5).iterrows():
    ax.annotate(gene, (row["log2FoldChange"], -np.log10(row["padj"])),
                fontsize=7, xytext=(4, 2), textcoords="offset points")

ax.set_xlabel("log₂ fold change (trt vs untrt)", fontsize=12)
ax.set_ylabel("−log₁₀ adjusted p-value", fontsize=12)
ax.set_title("airway dataset — DESeq2 via rosetta-bioc", fontsize=13)
ax.legend(frameon=False, fontsize=10)
plt.tight_layout()
plt.show()
```

---

## What just happened

| Step | Tool |
|------|------|
| Simulate RNA-seq counts | R `rnbinom` (500 genes × 6 samples) |
| Fit NB model + Wald test | DESeq2 1.52.0 (R 4.6.1 / Bioc 3.23) |
| Results back in Python | `rosetta.quick_deseq2()` → `RosettaDataFrame` |
| Volcano plot | matplotlib |

You wrote Python throughout. DESeq2 never left R — rosetta bridged the two
transparently.

### Next steps

```python
# LFC shrinkage
model = rb.DESeq2(counts_filt, metadata, "~ dex")
model.run_deseq()
shrunk = model.lfc_shrink(coef=2, type="apeglm")

# edgeR instead — same API
results_edger = rb.quick_edger(counts_filt, metadata, design="~ dex")
```
